Debugging by \textit{lastChange}

### 2010-10-27 trunk/papers/pldi2011/sigplanconf-template.tex
reword introduction

\section{Introduction}
Debugging is an inevitable part of programming, still hard and time-consuming. To fix a bug, developers have to reproduce and monitor the buggy execution several times to understand the program's unexpected behavior. Trial and error, guess-work and analyzing huge collected data are the inseparable parts of this process. According to \cite{LaToza}, developers spend about fifty percent of their time debugging. Therefore, enhancements to debugging techniques and tools can considerably save developers' time and improve programs' quality.

Locating defects cause the bug is the main part of debugging. A common strategy for locating defects is starting from bug symptoms and backward movement by following the origins to wrong values. While this strategy seems straightforward and effective, applying this strategy using traditional methods is complicated. There are two traditional approaches to debugging, breakpoint-based and log-based debugging. None of these approaches assist developer in finding origins to a wrong value. It is due to developer to search source files, find the list of possible origins to a wrong value and set breakpoints or insert log statements in a way that covers all possible origins.

The next dilemma is data collection. In breakpoint debugging,
developer has to memorize values or manually collect data at every
breakpoint hit. As the number of breakpoint hits increases, the
process of checking the program state, collecting data and resuming
the execution becomes cumbersome. While in breakpoint-based debugging,
the whole program state is available to developer, in log-based
debugging, developer has to decide about data should be collected when
inserts the log statement. It is very common that developer has to
repeat this step several times due to insufficient collected
data. Once the adequate data is collected, it still requires analyzing
and understanding. Developers usually end up in dealing with long log
files and analyzing huge collected data.

To address the mentioned problems, we propose a new functionality in
debuggers, \textit{lastChange}, which locates the origin of a wrong
value. It works based on program re-execution similar to the practice
in using breakpoints and log statements. Contrary to other
replay-based alogrithms which require the same re-executions
(instruction by instruction), the only prerequisite for
\textit{lastChange} functionality is bug reproducibility (i.e., there
is a test case which reproduces the bug). Assume that the program
execution is paused on a breakpoint and the developer is suspicious
about the value of an object property or a variable. The developer
queries \textit{lastChange} on the value and it is due to debugger to
locate and shows the last change of the object property or
variable. Debugger reproduces the buggy execution and collects limited
data and once the execution reaches the same place (i.e., the same
breakpoint hit), call it \textit{reproduction point}, it pauses the
execution, analyzes the collected data and shows the location of the
last change to the developer. The developer can also examine the
program state at the located point of execution, and continue
debugging by more \textit{lastChange} queries from that point.

>>>>

\section{Introduction}

According to \cite{LaToza}, developers spend about fifty percent of
their time debugging. To fix a bug, developers typically reproduce and
monitor the buggy execution several times to understand the program's
unexpected behavior. Trial and error, guess-work, and analyzing large
and complicated data make debugging difficult and time-consuming. Any
enhancement to debugging techniques and tools can save developers'
time, reducing development costs and improve programs' quality.

A common strategy for locating defects starts from bug symptoms and
works backwards, moving from a point in the program execution where a
value appears to be incorrect back to the point were that value was
set.  There are two conventional approaches used: breakpoint-based and
log-based debugging. Both approaches require tedious steps of
selecting data to be collected, collecting the data, then analyzing
the results.

In breakpoint debugging, developers select data to be collected by
searching through source files and setting breakpoints. To determine
where a value was set incorrectly, a developer must list and set
breakpoints at all possible points of the change for the value. At
every breakpoint the developer must determine if the location is in
fact related to the questionable value change then study the complex
debugger user interface and memorize values or manually collect
data. As the number of breakpoint hits increases, the process of
checking the program state, collecting data and resuming the execution
becomes cumbersome.

In log-based debugging, developers select data to be collected by
inserting statements for all points of possible change.  While in
breakpoint-based debugging, the whole program state is available to
developer, in log-based debugging, developer has to decide about data
should be collected when inserts the log statement. It is very common
that developer has to repeat this step several times due to
insufficient collected data, or to wait a long time because too much
data is recorded. Once the adequate data is collected, it still
requires analyzing and understanding. Developers usually end up in
dealing with long log files and analyzing huge collected data.  None
of these approaches assist developer in finding origins to a wrong
value.

To address these problems, we propose a new functionality in
debuggers, \textit{lastChange}, which locates the origin of a wrong
value. It uses program re-execution similar to the practice
in using breakpoints and log statements.

Imagine a program execution paused on a breakpoint and a developer
suspicious about the value of an object property or a variable. The
developer queries \textit{lastChange} on the value. The debugger
locates and shows the last change of the object property or
variable. Debugger reproduces the buggy execution and collects data
and once the execution reaches the same place (i.e., the same
breakpoint hit), it pauses the execution, analyzes the collected data
and shows the location of the last change to the developer. The
developer can also examine the program state at the located point of
execution, and continue debugging by more \textit{lastChange} queries
from that point.

We refer to the breakpoint used in \textit{lastChange} as the
\textit{reproduction point}: it is then point the developer knows
reproduces the bug and it is the point were the \textit{lastChange}
algorithm ends. The ability to identify this reproduction point, that
is a test case and a breakpoint after the bug is appears, is the only
prerequisite for \textit{lastChange} functionality.  This means
\textit{lastChange} can be implemented on top of a conventional
breakpoint debugger and it does not require a special environment to
create identical, instruction by instruction, re-executions.
